\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2018-10-15 paper/sp.tex
figure tweaks
references

This optimization is not convex---there are multiple optima in general.
In particular, there is a large, degenerate, pathological optimum where
the exponential function underflows, all parallaxes are predicted to vanish,
and there is no gradient of anything with respect to the parameter $D$-vector $\theta$.
This large, bad optimum (it is a local, not global, optimum) must be avoided in optimization.
In practice we avoid it by optimizing first for the very highest signal-to-noise
(in astrometric parallax) Training Set stars, and then use that first optimum as a
good first guess or initialization for the full optimization.
This method works because in the limit of high signal-to-noise, the problem asymptotically approaches
the convex problem of L1-regularized linear least-square fitting.
Optimization is performed with \code{scipy.optimize} using the
\acronym{L-BFGS-B} algorithm (HOGG CITE).

@@

The linear model permits straightforward propagation of uncertainty from the
feature inputs to the spectrophotometric parallaxes.
Although the model fundamentally assumes that the prediction features are noise-free
(see the ``good features'' assumption in \sectionname~\ref{sec:assumptions}),
there are in fact small uncertainties on the features that we can propagate to 
make uncertainty estimates on the spectrophotometric parallaxes.
\begin{eqnarray}
\ssigma_m^2 &\leftarrow& \sparallax_m^2\,\hat{\theta}^T\cdot C_m\cdot\hat{\theta}
\label{eq:unc}
\end{eqnarray}
where $\ssigma_m$ is the uncertainty on the inferred (or predicted)
spectrophotometric parallax $\sparallax_m$,
the left-hand side is a scalar product,
and $C_m$ is the (presumed diagonal, actually) covariance matrix of the input
features.
That is, $C_m$ is a diagonal matrix with the uncertainty variances of the elements
of $x_m$ along the diagonal.

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
HOGG CITE STASSUN MAYBE AND OTHERS?
The spectrophotometric parallaxes we generate
do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in the bottom panels of \figurename~\ref{fig:samples}.

@@

% HOGG: NEW FIGURE -- put some words in the text about the figure. 
\begin{figure}
\centering
\includegraphics[width=\textwidth]{./precision.pdf}
\caption{some smart words :) \label{fig:precision}}
\end{figure}

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from our model for stars that are
astrometrically confirmed members of known stellar clusters.
These clusters span some range in metallicity and age, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, these figures 
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlations
with parallax offsets.
\begin{figure}
\includegraphics[width=0.7\textwidth]{clusters.pdf}
\caption{The \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  four stellar clusters in the \apogee\ observing footprint. The spectrophotometric
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  are shown as vertical lines.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

\acknowledgements
It is a pleasure to thank
  Coryn Bailer-Jones (\acronym{MPIA}),
  Sven Buder (\acronym{MPIA}),
  Andy L. Jones,
  Daniel Michalik (\acronym{ESTEC}),
  Melissa Ness (Columbia),
  and the participants in the \acronym{MPIA} Milky Way Group Meeting
for help with this project.
This project was developed in part at the
2018 \acronym{NYC} Gaia Sprint, hosted by the Center for Computational Astrophysics of
the Flatiron Institute in New York City in 2018 June.

@@

\begin{thebibliography}{}
\bibitem[Abolfathi \etal(2018)]{dr14} Abolfathi, B., Aguado, D.~S., Aguilar, G., \etal\ 2018, \apjs, 235, 42 
\bibitem[Allende Prieto \etal(2008)]{aapogee} Allende Prieto, C., Majewski, S.~R., Schiavon, R., \etal\ 2008, Astronomische Nachrichten, 329, 1018 
\bibitem[Astropy Collaboration \etal(2013)]{astropy} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., \etal\ 2013, \aap, 558, A33 
\bibitem[Astropy Collaboration \etal(2018)]{astropy2} The Astropy Collaboration, Price-Whelan, A.~M., Sip{\H o}cz, B.~M., \etal\ 2018, arXiv:1801.02634 
\bibitem[Bailer-Jones \etal(2018)]{calj} Bailer-Jones, C.~A.~L., Rybizki, J., Fouesneau, M., Mantelet, G., \& Andrae, R.\ 2018, \aj, 156, 58 
\bibitem[Blanton \etal(2017)]{sdssiv} Blanton, M.~R., Bershady, M.~A., Abolfathi, B., \etal\ 2017, \aj, 154, 28 
\bibitem[Casey \etal(2016)]{casey} Casey, A.~R., Hogg, D.~W., Ness, M., \etal\ 2016, arXiv:1603.03040 
\bibitem[Gaia Collaboration \etal(2016)]{gaia} Gaia Collaboration, Prusti, T., de Bruijne, J.~H.~J., \etal\ 2016, \aap, 595, A1 
\bibitem[Gaia Collaboration \etal(2018)]{gaiadr2} Gaia Collaboration, Brown, A.~G.~A., Vallenari, A., \etal\ 2018, arXiv:1804.09365 
\bibitem[Garc{\'{\i}}a P{\'e}rez \etal(2016)]{aspcap} Garc{\'{\i}}a P{\'e}rez, A.~E., Allende Prieto, C., Holtzman, J.~A., \etal\ 2016, \aj, 151, 144 
\bibitem[Ho \etal(2017)]{ho} Ho, A.~Y.~Q., Ness, M.~K., Hogg, D.~W., \etal\ 2017, \apj, 836, 5 
\bibitem[Hogg(2018)]{gaialf} Hogg, D.~W.\ 2018, arXiv:1804.07766
\bibitem[Hunter(2007)]{matplotlib} Hunter, J.~D.\ 2007, CISE, 9(3), 90
\bibitem[Jones \etal(2001)]{scipy} Jones, E., Oliphant, E., Peterson, P. \etal\ 2001, \url{http://www.scipy.org/}
\bibitem[Kollmeier \etal(2017)]{sdssv} Kollmeier, J.~A., Zasowski, G., Rix, H.-W., \etal\ 2017, arXiv:1711.03234 
\bibitem[Lindegren \etal(2018)]{lindegren} Lindegren, L., Hernandez, J., Bombrun, A., \etal\ 2018, arXiv:1804.09366 
\bibitem[Lutz \& Kelker(1973)]{lk} Lutz, T.~E., \& Kelker, D.~H.\ 1973, \pasp, 85, 573 
\bibitem[Majewski \etal(2017)]{apogee} Majewski, S.~R., Schiavon, R.~P., Frinchaboy, P.~M., \etal\ 2017, \aj, 154, 94 
\bibitem[Marrese \etal(2018)]{xmatch} Marrase, P.~M. \etal\ 2018, in preparation
\bibitem[Martig \etal(2016)]{martig} Martig, M., Fouesneau, M., Rix, H.-W., \etal\ 2016, \mnras, 456, 3655 
\bibitem[Mora \etal(2018)]{gaiaarchive} Mora, A., Gonz{\'a}lez-N{\'u}{\~n}ez, J., Baines, D., \etal\ 2018, IAU Symposium, 330, 35 
\bibitem[Ness \etal(2015)]{cannon} Ness, M., Hogg, D.~W., Rix, H.-W., Ho, A.~Y.~Q., \& Zasowski, G.\ 2015, \apj, 808, 16 
\bibitem[Ness \etal(2016)]{nessage} Ness, M., Hogg, D.~W., Rix, H.-W., \etal\ 2016, \apj, 823, 114 
\bibitem[Ness \etal(2018)]{nessdopp} Ness, M., Rix, H.-W., Hogg, D.~W., \etal\ 2018, \apj, 853, 198 
\bibitem[Oliphant(2006)]{numpy} Oliphant, T.~E.\ 2006, \emph{A guide to NumPy}, Trelgol Publishing
\bibitem[P\'erez \& Granger(2007)]{ipython} P\'erez, F. \& Granger, B.~E.\ 2007, CISE, 9(3), 29
\bibitem[Skrutskie \etal(2006)]{zmass} Skrutskie, M.~F., Cutri, R.~M., Stiening, R., \etal\ 2006, \aj, 131, 1163 
\bibitem[Wilson \etal(2010)]{wapogee} Wilson, J.~C., Hearty, F., Skrutskie, M.~F., \etal\ 2010, \procspie, 7735, 77351C 
\bibitem[Wright \etal(2010)]{wise} Wright, E.~L., Eisenhardt, P.~R.~M., Mainzer, A.~K., \etal\ 2010, \aj, 140, 1868 
\end{thebibliography}

>>>>

This optimization is not convex---there are multiple optima in general.
In particular, there is a large, degenerate, pathological optimum where
the exponential function underflows, all parallaxes are predicted to vanish,
and there is no gradient of anything with respect to the parameter $D$-vector $\theta$.
This large, bad optimum (it is a local, not global, optimum) must be avoided in optimization.
In practice we avoid it by optimizing first for the very highest signal-to-noise
(in astrometric parallax) Training Set stars, and then use that first optimum as a
good first guess or initialization for the full optimization.
This method works because in the limit of high signal-to-noise, the problem asymptotically approaches
the convex problem of L1-regularized linear least-square fitting.
Optimization is performed with \code{scipy.optimize} using the
\acronym{L-BFGS-B} algorithm (\citealt{lbfgsb}).

@@

The linear model permits straightforward propagation of uncertainty from the
feature inputs to the spectrophotometric parallaxes.
Although the model fundamentally assumes that the prediction features are noise-free
(see the ``good features'' assumption in \sectionname~\ref{sec:assumptions}),
there are in fact small uncertainties on the features that we can propagate to 
make uncertainty estimates on the spectrophotometric parallaxes.
\begin{eqnarray}
\ssigma_m^2 &\leftarrow& \sparallax_m^2\,\hat{\theta}^T\cdot C_m\cdot\hat{\theta}
\label{eq:unc}
\end{eqnarray}
where $\ssigma_m$ is the uncertainty on the inferred (or predicted)
spectrophotometric parallax $\sparallax_m$,
the right-hand side is a scalar product,
and $C_m$ is the covariance matrix of the input
features.
In what follows we don't actually instantiate the full covariance matrix; we presume
that input features are independently measured;
that is, we make $C_m$ is a diagonal matrix with the uncertainty variances of the elements
of $x_m$ along the diagonal.

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
It also happens to be in the range of other values found in the literature (\citealt{arenou, zinn}).
Of course we really expect the offset to be a function of sky position and color (at least),
so the offset we find is not recommended for further use as any kind of universal value.
The spectrophotometric parallaxes we generate
do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in the bottom panels of \figurename~\ref{fig:samples}.

@@

% HOGG: NEW FIGURE -- put some words in the text about the figure. 
\begin{figure}
\centering
~\hfill\includegraphics[width=0.8\textwidth]{./precision.pdf}\hfill~
\caption{some smart words :) \label{fig:precision}}
\end{figure}

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from our model for stars that are
astrometrically confirmed members of known stellar clusters.
These clusters span some range in metallicity and age, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, these figures 
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlations
with parallax offsets.
\begin{figure}
~\hfill\includegraphics[width=0.7\textwidth]{clusters.pdf}\hfill~
\caption{The \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  four stellar clusters in the \apogee\ observing footprint. The spectrophotometric
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  are shown as vertical lines.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

\acknowledgements
It is a pleasure to thank
  Coryn Bailer-Jones (\acronym{MPIA}),
  Sven Buder (\acronym{MPIA}),
  Andy L. Jones,
  Daniel Michalik (\acronym{ESTEC}),
  Melissa Ness (Columbia),
  Joel Zinn (OSU),
  and the participants in the \acronym{MPIA} Milky Way Group Meeting
for help with this project.
This project was developed in part at the
2018 \acronym{NYC} Gaia Sprint, hosted by the Center for Computational Astrophysics of
the Flatiron Institute in New York City in 2018 June.

@@

\begin{thebibliography}{}
\bibitem[Abolfathi \etal(2018)]{dr14} Abolfathi, B., Aguado, D.~S., Aguilar, G., \etal\ 2018, \apjs, 235, 42 
\bibitem[Allende Prieto \etal(2008)]{aapogee} Allende Prieto, C., Majewski, S.~R., Schiavon, R., \etal\ 2008, Astronomische Nachrichten, 329, 1018 
\bibitem[Arenou \etal(2018)]{arenou} Arenou, F., Luri, X., Babusiaux, C., et al.\ 2018, \aap, 616, A17 
\bibitem[Astropy Collaboration \etal(2013)]{astropy} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., \etal\ 2013, \aap, 558, A33 
\bibitem[Astropy Collaboration \etal(2018)]{astropy2} The Astropy Collaboration, Price-Whelan, A.~M., Sip{\H o}cz, B.~M., \etal\ 2018, arXiv:1801.02634 
\bibitem[Bailer-Jones \etal(2018)]{calj} Bailer-Jones, C.~A.~L., Rybizki, J., Fouesneau, M., Mantelet, G., \& Andrae, R.\ 2018, \aj, 156, 58 
\bibitem[Blanton \etal(2017)]{sdssiv} Blanton, M.~R., Bershady, M.~A., Abolfathi, B., \etal\ 2017, \aj, 154, 28 
\bibitem[Byrd \etal(1995)]{lbfgsb} Byrd, R.~H., Lu,~P., Nocedal,~J., \& Zhu,~C. 1995, SIAM J. Sci. Comput., 16, 1190
\bibitem[Casey \etal(2016)]{casey} Casey, A.~R., Hogg, D.~W., Ness, M., \etal\ 2016, arXiv:1603.03040 
\bibitem[Gaia Collaboration \etal(2016)]{gaia} Gaia Collaboration, Prusti, T., de Bruijne, J.~H.~J., \etal\ 2016, \aap, 595, A1 
\bibitem[Gaia Collaboration \etal(2018)]{gaiadr2} Gaia Collaboration, Brown, A.~G.~A., Vallenari, A., \etal\ 2018, arXiv:1804.09365 
\bibitem[Garc{\'{\i}}a P{\'e}rez \etal(2016)]{aspcap} Garc{\'{\i}}a P{\'e}rez, A.~E., Allende Prieto, C., Holtzman, J.~A., \etal\ 2016, \aj, 151, 144 
\bibitem[Ho \etal(2017)]{ho} Ho, A.~Y.~Q., Ness, M.~K., Hogg, D.~W., \etal\ 2017, \apj, 836, 5 
\bibitem[Hogg(2018)]{gaialf} Hogg, D.~W.\ 2018, arXiv:1804.07766
\bibitem[Hunter(2007)]{matplotlib} Hunter, J.~D.\ 2007, CISE, 9(3), 90
\bibitem[Jones \etal(2001)]{scipy} Jones, E., Oliphant, E., Peterson, P. \etal\ 2001, \url{http://www.scipy.org/}
\bibitem[Kollmeier \etal(2017)]{sdssv} Kollmeier, J.~A., Zasowski, G., Rix, H.-W., \etal\ 2017, arXiv:1711.03234 
\bibitem[Lindegren \etal(2018)]{lindegren} Lindegren, L., Hernandez, J., Bombrun, A., \etal\ 2018, arXiv:1804.09366 
\bibitem[Lutz \& Kelker(1973)]{lk} Lutz, T.~E., \& Kelker, D.~H.\ 1973, \pasp, 85, 573 
\bibitem[Majewski \etal(2017)]{apogee} Majewski, S.~R., Schiavon, R.~P., Frinchaboy, P.~M., \etal\ 2017, \aj, 154, 94 
\bibitem[Marrese \etal(2018)]{xmatch} Marrase, P.~M. \etal\ 2018, in preparation
\bibitem[Martig \etal(2016)]{martig} Martig, M., Fouesneau, M., Rix, H.-W., \etal\ 2016, \mnras, 456, 3655 
\bibitem[Mora \etal(2018)]{gaiaarchive} Mora, A., Gonz{\'a}lez-N{\'u}{\~n}ez, J., Baines, D., \etal\ 2018, IAU Symposium, 330, 35 
\bibitem[Ness \etal(2015)]{cannon} Ness, M., Hogg, D.~W., Rix, H.-W., Ho, A.~Y.~Q., \& Zasowski, G.\ 2015, \apj, 808, 16 
\bibitem[Ness \etal(2016)]{nessage} Ness, M., Hogg, D.~W., Rix, H.-W., \etal\ 2016, \apj, 823, 114 
\bibitem[Ness \etal(2018)]{nessdopp} Ness, M., Rix, H.-W., Hogg, D.~W., \etal\ 2018, \apj, 853, 198 
\bibitem[Oliphant(2006)]{numpy} Oliphant, T.~E.\ 2006, \emph{A guide to NumPy}, Trelgol Publishing
\bibitem[P\'erez \& Granger(2007)]{ipython} P\'erez, F. \& Granger, B.~E.\ 2007, CISE, 9(3), 29
\bibitem[Skrutskie \etal(2006)]{zmass} Skrutskie, M.~F., Cutri, R.~M., Stiening, R., \etal\ 2006, \aj, 131, 1163 
\bibitem[Wilson \etal(2010)]{wapogee} Wilson, J.~C., Hearty, F., Skrutskie, M.~F., \etal\ 2010, \procspie, 7735, 77351C 
\bibitem[Wright \etal(2010)]{wise} Wright, E.~L., Eisenhardt, P.~R.~M., Mainzer, A.~K., \etal\ 2010, \aj, 140, 1868 
\bibitem[Zinn \etal(2018)]{zinn} Zinn, J.~C., Pinsonneault, M.~H., Huber, D., \& Stello, D.\ 2018, arXiv:1805.02650 
\end{thebibliography}
